---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mums2

<!-- badges: start -->
<!-- badges: end -->

mums2 is a package that was created to supply researchers with a collection of tools that they can use to analyze untargeted metabolomics data. This package will supply tools to analysis tandem Mass spectrometry (MS) data in a easy and efficient manner. The package supports importation and filtration of MS1 feature tables, annotations, de novo molecular formula predictions, scoring of MS2 spectra, clustering, alpha and beta diversity calculations, and community distance calculations.

## Installation

You can install the CRAN version with:

```{r, eval = FALSE}
install.packages("mums2")
```

You can install the development version of mpactr from [GitHub](https://github.com/mums2/mums2) with:

```{r, eval = FALSE}
devtools::install_github("mums2/mums2")
```

## Get started

See the [Getting Started](https://www.mums2.org/mpactr/articles/mpactr.html) page to get started.



## Getting help

If you encounter an issue, please file an issue on [GitHub](https://github.com/mums2/mums2/issues). Please include a minimal reproducible example with your issue.

## Contributing

Is there a feature you'd like to see included, please let us know! Pull requests are welcome on [GitHub](https://github.com/mums2/mums2/pulls). 
