---
title: "mums2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mums2}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


The mums2 package is designed to provide researchers with tools to analyze untargeted metabolomics data generated using tandeom mass spectroscopy from microbial communities. The overall approach taken to analyze metabolomics data parallels that used to analyze microbial communities using 16S rRNA gene sequencing data.


```{r setup}
library(mums2)
library(ggplot2)
library(mpactr)
```

## Import

Functions from the mpactR package are used to import data. mpactR allows us to import three different types of formats: Metaboscape, Progensis, and its own output. From there we can conduct a number of analysis on the data.

```{r import_data}
squid_data <- import_all_data(
  peak_table = mums2::mums2_example("squid_peak_table.csv"), 
  meta_data = mums2::mums2_example("squid_meta_data.csv"), 
  format = "None")
```

## Filter

After you import the data you can use from mpactR to filter the data. There are four different filters included in mpactR: `check_mispicked_ions()`, `filter_group()`, `filter_cv()`, and `filter_insource_ions()` (You can find more information on mpactR's website).

```{r  R.options=list(max.print=15)}
squid_filter <- squid_data |>
  filter_peak_table(filter_mispicked_ions_parameters()) |>
  filter_peak_table(filter_cv_parameters(cv_threshold = 0.2, 
                                         cv_param = "mean")) |>
  filter_peak_table(filter_group_parameters(group_threshold = 0.1,
                                            "Blanks")) |>
  filter_peak_table(filter_insource_ions_parameters())

squid_filter
```

## Convert from rt to "RT in minutes" or "rt in seconds"

Sometimes a MS2 file will report the retention time in minutes but the MS1 file will use seconds. This mismatch will cause the MS1 data to match incorrect peaks in the MS2 data. The `change_rt_to_seconds_or_minutes()` function allows users to synthesize data with different units. Be aware that this function modifes the mpactr object in place. Therefore, you will need to call the function again to revert the units.
```{r R.options=list(max.print=15)}
get_peak_table(squid_filter)$rt
```

The retention time is currently in minutes, however, some mgf files may display retention time in seconds. To ensure that our MS1 data will proper match our MS2 data, we can convert from minutes to seconds. 

```{r R.options=list(max.print=15)}
change_rt_to_seconds_or_minutes(squid_filter, "seconds")
change_rt_to_seconds_or_minutes(squid_filter, "minutes")
```

***Which command here do the users need to carry out the analysis?***

## Preprocess MS2 data

Once the mpactr_object is created one can use a mgf/mzxml/mzml to match MS1 and MS2 peaks. The `ms2_ms1_compare()` function reads a list of mgf files and matches them with a MS1 spectra based on the mass-charge ratio and retention time tolerance. If there are multiple peaks, it will select the peak with the highest intensity. Keep in mind that MS2 spectra files are very memory intensive, they can be anywhere from 1 MB to 100 GB. Therefore, depending on how big the file is, it might take a few moments for the function to complete.

```{r  R.options=list(max.print=15)}
matched_data <- ms2_ms1_compare(
  mums2_example("12152023_Coculture_with_new_JC1.gnps.mgf"), squid_filter, 2, 6)

matched_data
```

## Chemical formula prediction

To add annotation information to the data, mums2 will generate *de novo* chemical formula predictions using fragmentation trees. The `compute_molecular_formulas()` function uses MS2 data to generate the most explained chemical formula and return it as a result (for more information: [Fragmentation Trees](https://doi.org/10.1093/bioinformatics/btn270)).

```{r  R.options=list(max.print=15)}
matched_data <- compute_molecular_formulas(matched_data, 3, 16)
matched_data$predicted_molecular_formulas
```

## Annotations

Beyond predicting the molecular formula, it is possible to use the `annotate_ms()` function to annotate the data in the matched_ms1_ms2 object. By using an msp file as a reference and using a specified scoring algorithm we can compute similarity between peaks in the msp file and generate annotations using the cosine squared or another calculator. 

```{r  R.options=list(max.print=15)}
psu_msmls <- read_msp(mums2_example("PSU-MSMLS.msp"))
annotations <- annotate_ms2(
  mass_data = matched_data, reference = psu_msmls,
  scoring_params = gnps_params(0.5), precursor_tolerance = 1000,
  min_score =  0.1, chemical_min_score = .1)

annotations
```

## Scoring/Distance

We have implemented distance calculation approaches to generate distances between all pairs of compounds. We can use that distance to cluster the data based on their similarity.

```{r  R.options=list(max.print=15)}
dist <- dist_ms2(
  data = matched_data, cutoff = 0.3, precursor_thresh = 2,
  score_params = gnps_params(0.5), min_peaks = 0)

dist
```

## Operational Metabolomic Units (OMUs)

Once a distance matrix is generated between metabolites, they can be clustered into OMUs using functions from the clustur package

```{r  R.options=list(max.print=15)}
cluster_results <- cluster_data(
  distance_df = dist, ms2_match_data = matched_data, cutoff = 0.3,
  cluster_method = "opticlust")

cluster_results
```

## Clusters and Annotations

After you have clustered your data into OMUs the consensus annotation for the spectra within the OMU can be determined usin ghte annotate_ms2() function.

```{r}
annotations <- annotate_ms2(matched_data, psu_msmls, gnps_params(0.5), 
                            1000, 0.1, .1, cluster_results)

annotations
```

## Create the community object/matrix

Now that the data have been clustered into OMUs, it is possible to use the abundance of each OMU across samples to perform ecological analyses. First, it is necessary to transform it using the `create_community_object()` function.

```{r  R.options=list(max.print=15)}
community_w_omus <- create_community_matrix_object(cluster_results)
get_community_matrix(community_w_omus)
```

If you decide that you would rather work with distinct MS data rather than OMUs, the following will generate a community matrix using the individual MS data.

```{r}
community_wo_omus <- create_community_matrix_object(matched_data)
get_community_matrix(community_wo_omus)
```

## Run the diversity metrics


<!-- The rarefaction section is unnecessary as written...
### Rarefaction -->

<!-- Once you create your community object, we can finally compute the diversity metrics on our data. Below this you can see the `rarefaction()` function. -->

<!-- ```{r  R.options=list(max.print=15)} -->
<!-- rarefy_new <- rarefy_ms( -->
<!--   community_object = community_obj, size = 4000, threshold = 100) -->

<!-- rarefy_new[1:10, 1:10] -->
<!-- ``` -->

### Alpha Diversity

You can also compute the alpha diversity using the Shannon and Simpson diversity indices.

```{r  R.options=list(max.print=15)}
diversity(community_object = community_w_omus, diversity_index = "shannon")
diversity(community_object = community_w_omus, diversity_index = "simpson")
```

### Beta Diversity

Beta diversity values can be calculated using the Bray-Curtis, Theta-YC, Jaccard, Hamming, Moristia-Horn, and Sorenson distance calculators. Users are strongly encouraged to use rarefaction to control for uneven sampling effort when calculating any diversity metric. This function call calculates Bray-Curtis distances without using rarefaction

```{r}
diversity(community_object = community_w_omus, diversity_index = "bray")
```

Here are examples of using rarefaction to calculate Bray-Curtis distances using OMU or individual metabolites

```{r  R.options=list(max.print=15)}
bray_w_omus <- dist_shared(community_object = community_w_omus, size = 4000,
                          threshold = 100, diversity_index = "bray")
bray_w_omus

bray_no_omus <- dist_shared(community_object = community_wo_omus, size = 4000,
                           threshold = 100, diversity_index = "bray")
bray_no_omus
```

## Visualizations

### PCOA

Princpal Coordinate Analysis (PCoA) is a statistical technique used to show similarities between samples based on the beta diversity values. There are a few steps needed to perform PCoA in R. To start, a special distance matrix object is needed. Using `community_object_to_distance_object()` function, we can transform the distances that were generated from the `dist_shared()` function (e.g., `bray_w_omus`). Once a distance matrix is created, we can run PCoA using the `cmdscale()` function. Once the result is computed, we clean up the data and plot.

```{r, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
dist_obj <- community_object_to_distance_object(bray_w_omus)

pcoa <- cmdscale(dist_obj, k = 2, eig = T, add = T)
pcoa_points <- as.data.frame(pcoa$points)
variance <- round(pcoa$eig*100/sum(pcoa$eig), 1)
names(pcoa_points) <- paste0("PCOA", seq(1:2))
pcoa_points$Samples <- rownames(pcoa_points)

meta_data <- get_meta_data(squid_filter)
for(i in 1:nrow(pcoa_points)) {
  idx <- which(pcoa_points$Samples[[i]] == meta_data$Injection)
  pcoa_points$Samples[[i]] <- meta_data$Biological_Group[[i]]
  
}

pcoa_points |>
  ggplot(aes(x=PCOA1, y = PCOA2, color = Samples)) +
    geom_point() +
    theme(
      legend.text = element_text(size = 10), 
      legend.title = element_text(size = 15)
    )
```


### Hierarchical Clustering

Another visualization approach is to perform hierarchical clustering using the hclust() function. Hierarchical clustering generates a tree of samples based on their beta diversity values to show the similarities of the samples

```{r, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
hclust_result <- hclust(dist_obj, "complete")
par(cex=0.6, mar=c(5, 8, 4, 1))
plot(hclust_result)
```

