---
title: "mums2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mums2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mums2)
library(ggplot2)
library(mpactr)
```



# Import
To import data inside of mums2, we are using the mpactR package. mpactR allows us to import three different types of formats: Metaboscape, Progensis,
and its own output. From there we can conduct a number of analysis on the data.
```{r import_data}

squid_data <- import_all_data(peak_table = mums2::example("squid_peak_table.csv"), 
                              meta_data = mums2::example("squid_meta_data.csv"), 
                              format = "None")

```


# Filter
After you import the data we are able to filter the data using filters including in mpactR. There are four different filters included in mpactR, 
`check_mispicked_ions()`, `filter_group()`, `filter_cv()`, and `filter_insource_ions()` (You can find more information on mpactR's website).
```{r  R.options=list(max.print=15)}
  squid_filter <- squid_data |>
    filter_peak_table(filter_mispicked_ions_parameters()) |>
    filter_peak_table(filter_cv_parameters(cv_threshold = 0.2, cv_param = "mean")) |>
    filter_peak_table(filter_group_parameters(group_threshold = 0.1, "Blanks")) |>
    filter_peak_table(filter_insource_ions_parameters())
  squid_filter
```


# Preprocess ms2 data
Once you create your mpactr_object, you can use a mgf/mzxml/mzml to match ms1 and ms2 peaks. Our `ms2_ms1_compare()` function reads a list of mgf files and matches them with a ms1 spectra based on the mass-charge ratio and retention time tolerance. If there are multiple peaks, it will select the peak with the highest intensity. Keep in mind that ms2 spectra files are very memory intensive, they can be anywhere from 1mb to 100gbs. Therefore, depending on how big your file is, do not expect this function to be overly fast.  
```{r  R.options=list(max.print=15)}
  matched_data <- ms2_ms1_compare(example("12152023_Coculture_with_new_JC1.gnps.mgf"), squid_filter, 2, 6)
  matched_data
  
```


# Annotations
We are also able annotate our matched_ms1_ms2 object. By using an msp file as a reference and using a specified scoring algorithm(whether it be a cosine squared algorith or anything else) we can compute similairty between peaks in the msp file and generate annotations. 
```{r  R.options=list(max.print=15)}


  psu_msmls <- read_msp(example("PSU-MSMLS.msp"))
 
  annotations <- annotate_ms2(query = matched_data, reference = psu_msmls,
                              score_params = gnps_params(0.5), precursor_tolerance =  2, min_score = .2, min_peaks = 0)
```

# Scoring/Distance
We have implemented a scoring and/or distance algorithm that allows you to generate a score between two different features. We can use that score to cluster the data based on closeness.
```{r  R.options=list(max.print=15)}
 dist <- dist_ms2(data = matched_data, cutoff = 0.3, precursor_thresh = 2, score_params = gnps_params(0.5), min_peaks = 0)
 dist
```

# OMUs (Clustering)
Once we have generated a distance matrix, we can use that matrix to cluster our data using the clustur package (more information can be found on the clustur website). clustur is an R package that contains the clusturing algorithms from mothur (a software that is used to anaylsis microbial ecology).
```{r  R.options=list(max.print=15)}

  cluster_results <- cluster_data(distance_df = dist, ms2_match_data = matched_data, cutoff = 0.3, cluster_method = "opticlust")
  cluster_results

```

# Clusters and Annotations
After you have clustered your data together, you can add the OMU the feature resides in to your annotation table for a more accurate data frame.

```{r}
 annotations <- annotate_ms2(query = matched_data, reference = psu_msmls,
                              score_params = gnps_params(0.5), precursor_tolerance =  2, min_score = .2, min_peaks = 0, cluster_results)

```

# Create the community object/matrix
Once you clustered your data, you have to transform it into a data-usable format. We have supplied the `create_community_object()` function for this purpose.
```{r  R.options=list(max.print=15)}
  # Create community object
  community_obj <- create_community_matrix_object(cluster_results)
  mat <- get_community_matrix(community_obj)

  # You can also do this without clustering
  community_obj_no_cluster <- create_community_matrix_object(matched_data)
  mat2 <- get_community_matrix(community_obj_no_cluster)
  mat[, 1:10]
```

# Run the diversity metrics
## Rarefaction
Once you create your community object, we can finally compute the diversity metrics on our data. Below this you can see the `rarefaction()` function.
```{r  R.options=list(max.print=15)}
  # Rarefacton
  rarefy_new <- rarefy_ms(community_object = community_obj, size = 4000, threshold = 100)
  rarefy_new[1:10, 1:10]
```


## Alpha Diversity
You can also compute the alpha diversity. We currently only support shannon, and simpsons diversity.
```{r  R.options=list(max.print=15)}
  # Calculate Alpha Diversity
  shannon_diversity <- diversity(community_object = community_obj, diversity_index = "shannon")
  simpson_diversity <- diversity(community_object = community_obj, diversity_index = "simpson")
  shannon_diversity
  simpson_diversity

```

## Beta Diversity
Beta diversity is also included inside of this package, included diversity metrtics: bray, thetayc (Yue and Clayton), jaccard distance, hamming distance, moristia horn distance, and soren distance. 
```{r}
  #Calculate Beta Diversity
  braycurits <- diversity(community_object = community_obj, diversity_index = "bray")
  braycurits[1,]
```

## Average Subsampled Dissimilarity
Finally you can average the diversity metrics using the `average_subsampled_diversity()` function. This allows you to run the rarefaction and diversity command n number of times. This will generate a average distance matrix. This allows you to ensure that your rare species were included inside of the analysis. 
```{r  R.options=list(max.print=15)}
  # Avg dist
  avg <- dist_shared(community_object = community_obj, size = 4000, threshold = 100, diversity_index = "bray", iterations = 100)
  # Avg dist without clustering
  avg_no_cluster <- dist_shared(community_object = community_obj_no_cluster, size = 4000, threshold = 100, diversity_index = "bray", iterations = 100)
  avg
```


# Visualizations
## PCOA
PCoA or Princple coordinate analysis is a statistical technique used to show similiarities or disimilarities in your data. PCA or Principle component analysis uses dimension reduction techniques to look for similiarities or variance inside data. These two method perform similiar metric, but PCA can only be performed on euclidean distances. PCOA can be performed using a multitude of distance calculations (such as bray curtis or jaccard distance). In ecology, because there are numerous ways to calculate diversity between species, pcoa allows us to visualize differences between samples.

There are a few steps needed to calculate a pcoa in R. To start, you have to create a dist object or distance matrix. Using our `community_object_to_distance_object()` function, we can transform your distances that were generated from the `dist_shared()` function into a distance matrix. Once your distance matrix is created, we can conduct a pcoa or Classical Multidimensional Scaling(MDS) using the `cmdscale()` function. Once the result is computed, we clean up the data and plot.
```{r, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
dist_obj <- community_object_to_distance_object(avg)

pcoa <- cmdscale(dist_obj, k = 2, eig = T, add = T)
pcoa_points <- as.data.frame(pcoa$points)
variance <- round(pcoa$eig*100/sum(pcoa$eig), 1)
names(pcoa_points) <- paste0("PCOA", seq(1:2))
pcoa_points$Samples <- rownames(pcoa_points)

meta_data <- get_meta_data(squid_filter)
for(i in 1:nrow(pcoa_points)) {
  idx <- which(pcoa_points$Samples[[i]] == meta_data$Injection)
  pcoa_points$Samples[[i]] <- meta_data$Biological_Group[[i]]
  
}


pcoa_points |>
  ggplot(aes(x=PCOA1, y = PCOA2, color = Samples)) +
    geom_point() +
    theme(
      legend.text = element_text(size = 10), 
      legend.title = element_text(size = 15)
    )

```


## Hierarchical Clustering

Another way to visualize your data is by using hclust or Hierarchical Clustering. Hierarchical clustering generates a tree of distances to show the similarities between the samples.
```{r, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
hclust_result <- hclust(dist_obj, "complete")
par(cex=0.6, mar=c(5, 8, 4, 1))
plot(hclust_result)
```