---
title: "mums2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mums2}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The mums2 package is designed to provide researchers with tools to analyze untargeted metabolomics data generated using tandem mass spectroscopy from microbial communities. The overall approach taken to analyze metabolomics data parallels that used to analyze microbial communities using 16S rRNA gene sequencing data.

```{r setup}
library(mums2)
library(ggplot2)
library(mpactr)
library(networkD3)
library(reshape2)
```

## Process data

Before we begin to analyze your data, we have to process it into a readable `data.frame` or object that we can view and transform. To process your data we need to run a two functions: `import_all_data()`, and `ms2_ms1_compare()`. `import_all_data()` creates an object reflecting MS1 data, and `ms2_ms1_compare()` assigns MS2 spectra to your MS1 data. At this point, you may need to filter out noise, or transform it before you can properly analyze it. To accommodate for those issues, we have two other functions: `filter_peak_table()`, `change_rt_to_seconds_or_minute()`. `filter_peak_table()` allows your to remove low-quality features and `change_rt_to_seconds_or_minute()` allows you to transform your retention time to minutes or seconds. This allows you to ensure your MS1 data retention time matches your MS2 data retention time. Below will explain in more detail what each function does and how to go through the pipeline.

### Import

The `import_all_data()` function takes in a peak_table and meta_data tables as input and converts them into a `mpactr` object (its a wrapper for `mpactr::import_data()`). The format parameter accepts three different formats: Metaboscape, Progenesis, and None. You can read up more about each of the formats here(`mpactr::import_data()`). If you need a deeper understanding of what format the peak_table and meta_data file should be in, take a look at mpactr's [getting started page](https://www.mums2.org/mpactr/articles/mpactr.html).

Meta_data is a csv that provides information about the samples. In order for your meta_data to be valid, it needs the following columns: "Injection", "Sample_Code", " Biological_Group." The "Injection" column is the name of the samples, it should match the samples inside of the peak_table. "Sample_Code" is the id for you technical replicates. Finally, "Biological_Group" is another id, but for your biological replicate groups (you can learn more here `mpactr::import_data()`).

```{r import_data}
data <- import_all_data(
  peak_table = mums2::mums2_example("230112_botryllus_peaktable.csv"), 
  meta_data = mums2::mums2_example("meta_data_boryillus.csv"), 
  format = "Progenesis")
```


#### Peak Table

Below is the expected format for a progenesis peak table. It contains samples as columns and features as rows. The feature intensities are expected to be un-normalized. 

```{r peak_table, R.options=list(max.print=40)}
read.csv(mums2::mums2_example("230112_botryllus_peaktable.csv"), check.names = FALSE)
```

#### Metadata

The expected format for metadata is below. The metadata file needs to contain at minimum columns for "Injection", "Sample_Code", and " Biological_Group." 
```{r metadata, R.options=list(max.print=40)}
read.csv(mums2::mums2_example("meta_data_boryillus.csv"), check.names = FALSE)
```

### Filter

After importing the data, you can use functions from mpactR to filter the data. There are four different filters included in mpactR: `mpactr::filter_mispicked_ions()`, `mpactr::filter_group()`, `mpactr::filter_cv()`, and `mpactr::filter_insource_ions()` (You can find more information on [mpactR's](https://www.mums2.org/mpactr/) website). Although data fitlering is not required, it will help reduce noise and correct peak selection errors, which will also speed up the analysis.

```{r filter_data}
filtered_data <- data |>
  filter_peak_table(filter_mispicked_ions_params()) |>
  filter_peak_table(filter_cv_params(cv_threshold = 0.2)) |>
  filter_peak_table(filter_group_params(group_threshold = 0.1,
                                            "Blanks")) |>
  filter_peak_table(filter_insource_ions_params())

head(get_peak_table(filtered_data), 3)
```

### Convert from rt to "rt in minutes" or "rt in seconds"

Sometimes a MS2 file will report the retention time in minutes but the MS1 file will use seconds. This mismatch will cause the MS1 data to match incorrect peaks in the MS2 data. The `change_rt_to_seconds_or_minute()` function allows users to synthesize data with different units. Be aware that this function modifies the mpactr object in place. Therefore, you will need to call the function again to revert the units. Below will display a vector of retention time.

```{r show_rt_of_filtered_data, R.options=list(max.print=15)}
get_peak_table(filtered_data)$rt
```

```{r convert_rt_to_minutes_or_seconds}

filtered_data <- change_rt_to_seconds_or_minute(mpactr_object = filtered_data, rt_type = "seconds")
head(get_peak_table(filtered_data), 1)

filtered_data <- change_rt_to_seconds_or_minute(mpactr_object = filtered_data, rt_type = "minutes")
head(get_peak_table(filtered_data), 1)
```

### Preprocess MS2 data

Using the generated mpactr_object from calling the `import_all_data()` function we can use a .mgf/.mzxml/.mzml file to match MS1 and MS2 peaks. The `ms2_ms1_compare()` function reads a list of mgf files and matches them with a MS1 spectra based on the mass-charge ratio and retention time tolerance. If there are multiple matches, it will select the MS2 spectra with the highest intensity. Keep in mind that MS2 spectra files are very memory intensive, they can be anywhere from 1 MB to 100 GB. Therefore, depending on how big the file is, it might take a few moments for the function to complete.

`ms2_ms1_compare()` generates a list of data with two data.frames ("ms1_data", "peak_data"), a list ("peak_data"), and a character vector ("samples").

**ms2_matches** - One of the two data.frames, "ms2_matches", is a data.frame that contains five columns: "mz", "rt", "ms1_compound_id", "spectra_index", and "ms2_spectrum_id." "mz" and "rt", represent the MS2 mass to charge ratio and retention time. "ms1_compound_id" represents your MS1 compound id that was imported from the feature table. "spectra_index" matches the MS2 data with its MS2 spectrum. Finally, "ms2_spectrum_id" similar to the "ms1_compound_id", is the generated name to represent your MS2 spectra (the name is a combination of your mz and rt). This is necessary to properly compare compounds. Since compounds with similiar structures are expected to have similar MS2 patterns, we can use algorithmic techniques to compute the similarity between two spectra. This allows us to generate annotations and cluster similar spectra (MS2 matched information) together.

**ms1_data** - The other "data.frame", "ms1_data", is a data.frame of the created mpactr object.

**peak_data** - The list that is generate from `ms2_ms1_compare()` is named "peak_data." "peak_data" is a collection of MS2 peak list. A peak list a collection of fragment ions, they all have a value to represent their intensity and mass-charge ratio.

**samples** - The last slot inside of the list is a character vector named "samples." This is a list of the groups/samples contained inside of your peak_table file.

```{r preprocess_ms2_data}
matched_data <- ms2_ms1_compare(
  ms2_files = mums2_example("botryllus_v2.gnps.mgf"),mpactr_object = filtered_data, mz_tolerance = 5, rt_tolerance = 6)

head(matched_data$ms2_matches)

head(matched_data$ms1_data, 1)

matched_data$peak_data[[1]]

matched_data$samples
```

## Generate Metadata

Once you preprocess your data, we can start to generate additional metadata like molecular formulas and annotations. Using the `compute_molecular_formulas()` function, we can generate molecular formulas and `annotate_ms2()` allows us to annotate our data based on reference databases. This allows us to annotate additional data to unknown features or confirm known features. Below will explain in further detail how these functions can be used.

### Molecular formula prediction

To add annotation information to the data, mums2 will generate *de novo* molecular formula predictions using fragmentation trees. The `compute_molecular_formulas()` function uses MS2 data to generate the most explained molecular formula and return it as a result (for more information: [Fragmentation Trees](https://doi.org/10.1093/bioinformatics/btn270)). The most explained molecular formula simply means the molecular formula that is explained by the most peaks in the compound. In order to create a fragmentation tree we have to generate every possible molecular formula the parent mass can create (using only CHNOPS). We call these formulas the candidate formulas, formulas that have the potential to be the predicted formulas. We then look at every mass and intensity pair inside of the compound and compute every molecular formula. We then create a one directional graph based on all the molecular formulas using. A molecular formula will point to another if it is a subformula of another formula (meaning it contains at most as many atoms as the parent formula). Finally, we can compute a score for each one of the nodes uses methods like Ring Double Bond equivalents, compute molecular formula score, etc. You can learn how other open-sourced software such as [MZMine](https://mzio.io/mzmine-news/) and [Sirius](https://github.com/sirius-ms/sirius) generate molecular formula. Due to the time this function will take to run, we are going to use a small testset.

It is possible for a formula to be unable to be generated. In this case we return a NA to represent an unknown molecular formula.

Warning messages will be printed if no molecular formula is generated or there is only one possible molecular formula. 

```{r  chemical_formula_prediction, R.options=list(max.print=15)}

data_small <- import_all_data(
  peak_table = mums2::mums2_example("botryllus_pt_small.csv"), 
  meta_data = mums2::mums2_example("meta_data_boryillus.csv"), 
  format = "None") |> 
    filter_peak_table(filter_mispicked_ions_params()) |>
    filter_peak_table(filter_cv_params(cv_threshold = 0.05)) |>
    filter_peak_table(filter_group_params(group_threshold = 0.1,
                                              "Blanks")) |>
    filter_peak_table(filter_insource_ions_params())

matched_data_small <- ms2_ms1_compare(
  ms2_files = mums2_example("botryllus_v2.gnps.mgf"), mpactr_object = data_small, mz_tolerance = .5, rt_tolerance = 6)


matched_data_small <- compute_molecular_formulas(mass_data = matched_data_small, parent_ppm = 3)
matched_data_small$predicted_molecular_formulas
```

### Annotations

Beyond predicting the molecular formula, it is possible to use the `annotate_ms2()` function to annotate the data in the matched_ms2_ms1 object using reference databases. A reference database can be input as msp files using the `read_msp()` function. It returns a reference database that can be used as an input for the `annotate_ms2()` function. In mum2 package, MS2 spectral similarity can be determined using either spectral entropy (for more information: [Spectral Entropy](https://doi.org/10.1038/s41592-021-01331-z)) or gnps algorithm (for more information: [GNPS](https://doi.org/10.1038/nbt.3597)). While gnps uses an modified cosine score to compute similarity between spectra, spectral entropy takes advantage of the total intensities of the spectra. We determine which method we use by supplying either, `gnps_param()` or `spec_entropy_params()`. Using these two methods we are able to effectively generate a collection of annotations based on the reference databases. We have a small massbank database provided from [MSDial](https://systemsomicslab.github.io/compms/msdial/main.html#MSP) that is included in the package and can be used to annotate data.

```{r annotations}
reference_db <- read_msp(msp_file = mums2_example("massbank_example_data.msp"))
annotations <- annotate_ms2(
  mass_data = matched_data, reference = reference_db,
  scoring_params = modified_cosine_params(0.5), ppm = 1000,
  min_score =  0.1, chemical_min_score = 0)

annotations[1:5,]
```

## Create OMUs

Let's look into generating OMUs. OMUs are cluster of metabolites with similar MS2 spectral pattern and can be used for numerous analysis. To properly cluster your data together, you need to generate some similarity or distance between the features of your data. This is where our `dist_ms2()` function comes in. After you generate a distance `data.frame` we can use the `cluster_data()` function to create your OMUs. Below will show you the process how this works.

### Scoring/Distance

We have implemented two different distance calculations to generate distances between compounds. To generate the distances you can use the gnps algorithm (`modified_cosine_params()`) or spectral entropy algorithm (`spec_entropy_params()`). Just like above, being able to compute the similarity between MS2 spectra is what allows us to cluster data. We can also use the similarity distances to generate a simple distance data.frame for later use.

```{r  scoring_distance, R.options=list(max.print=15)}
dist <- dist_ms2(
  data = matched_data, cutoff = 0.3, precursor_threshold = -1,
  score_params = modified_cosine_params(0.5), min_peaks = 0)
dist
```

### Operational Metabolomic Units (OMUs)

OMUs or Operational Metabolomic Units are clusters of metabolites. We cluster metabolites based on how similar their ms2s are to each other. The `dist_ms2()` function generates a distance `data.frame`, or a `data.frame` of closeness between metabolites. Using the generated object along with the matched_ms2_ms1 object, we are able to cluster the data together to generate OMUs using the `cluster_data()` function. This function returns a list with five slots; label, abundance, cluster, cluster_metrics, and iteration metrics.

**label** - Represents the cutoff distance used for the cluster. **abundance** - A `data.frame()` that shows the absolute abundance of the clusters by sample. **cluster** - A `data.frame()` that shows which features clustered together **cluster_metrics** - A `data.frame()` representing metrics for how the clusters were formed. **iteration_metrics** - A `data.frame()` that shows how the data was clustered at each iteration.

```{r  clustering, R.options=list(max.print=15)}
cluster_results <- cluster_data(
  distance_df = dist, ms2_match_data = matched_data, cutoff = 0.3,
  cluster_method = "opticlust")

cluster_results
```

### Display OMUs when you annotate

After your data has been clustered, if you wish to see which OMUs the annotated features are in, you can supply that to the annotation function using the "cluster_data" parameter. Doing so will add a column named OMU. This column will display the OMU the feature is represented in.

```{r omu_annotations}
annotations <- annotate_ms2(mass_data = matched_data, reference = reference_db, scoring_params = modified_cosine_params(0.5), 
                            ppm = 1000, min_score = 0.7, chemical_min_score = 0, cluster_data = cluster_results)

annotations[1:5,]
```

## Ecological Analysis

After all of your data is processed, we can finally start analysis your data. To begin, you have to create a "community_object". A community object is a Rcpp Object used for all ecological analysis. You can generate a "community_object" by running the `create_community_matrix_object()`. Once your object is created, we can start computing the diversity inside of the samples. We can compute beta (`dist_shared()`) and alpha diversity (`alpha_summary()`). For more information and how to analysis beta and alpha diversity, or how to further analysis your data, continue reading below.

### Create the community object/matrix

Now that the data have been clustered into OMUs, it is possible to use the abundance of each OMU across samples to perform ecological analyses. First, it is necessary to transform it using the `create_community_matrix_object()` function. This will create a Rcpp object that contains all of your generated ecological data. If you wish to see what your data looks like, you can call the `get_community_matrix()` function.

Packages like vegan opt for using a `matrix` object as input for ecological analyses (see more `vegan::rarefy()`). However, if we were to use the same methodology, the computation would take far too long. In particular, the rarefaction algorithm. Rarefaction is the process of randomly selecting n number of samples without replacement. With n being the user defined sum threshold. However, in mass spectrometry, in order to account for machine limits and background noise, we have to define a sample threshold. So, rarefaction is the process of randomly selecting n number of samples without replacement, with n being the user defined sum threshold. **After selecting n number of samples, we only add the samples that are above the size threshold. If the total is below the sum threshold, we continue until we are equal to or above the sum threshold.** Therefore, adding the size threshold adds another layer complexity to the algorithm. In order to retain efficiency, we create a custom object that contains all of the necessary data for the calculation.

```{r  community_object, R.options=list(max.print=15)}
community_w_omus <- create_community_matrix_object(data = cluster_results)
community_w_omus
```

You have the option of not clustering your data to generate a community object as well. However, by not creating OMUs, it may be difficult to determine unknown metabolites. When creating OMUs, all alike data is clustered together to create an OMU, which means similar species stay together.

```{r community_object_no_omus, R.options=list(max.print=30)}
community_wo_omus <- create_community_matrix_object(data = matched_data)
community_wo_omus
```

### Community Matrix

The vegan package uses an object called a "community matrix." A community matrix is a `matrix` that represents samples as rows, and species as columns. Vegan does not account for machine limits and background noise, so there is no threshold needed when vegan rarefies data. In are effort to make a fast and efficient tool, we have opted to create a `community object`. It is the same as a community matrix, but it is created and held in a Rcpp class for speedy access. However, if you wish to use vegan for other analysis, we have created a `get_community_matrix()` function. This function will convert your community object into a community matrix for ease of use.

```{r community_matrix, R.options=list(max.print=30)}
  get_community_matrix(community_object = community_w_omus)
```

### Alpha Diversity

Using your created community matrix object, you are able to compute alpha diversity. Alpha diversity represents diversity (or richness) on a local scale. In our case, it represents the diversity inside of a single sample. You can determine alpha diversity in this package using two different methods: Simpson's and Shannon's diversity. Simpson's diversity represents the diversity on a scale of zero to one, zero meaning no diversity and one meaning maximum diversity. Shannon's diversity is not bound by zero, essentially, the higher the value, the more diversity there is.

```{r  alpha_summary, R.options=list(max.print=15)}
alpha_summary(community_object = community_w_omus, size = 4000, threshold = 100,
              diversity_index = "simpson", subsample = TRUE)
```

### Beta Diversity

Beta diversity values can be calculated using the Bray-Curtis, Theta-YC, Jaccard, Hamming, Moristia-Horn, and Sorenson distance calculators. Users are strongly encouraged to use rarefaction to control for uneven sampling effort when calculating any diversity metric. By default, the `dist_shared()` function uses subsample = TRUE, which automatically rarefies the samples.

Beta diversity represents the dissimilarity of two different samples. With this calculation the user can generate distances between the community that can be used for visualizations. And, with all of the annotations, and clusters that were computed earlier, you can see just how dissimilar/similar the samples are.

Below are examples of using rarefaction to calculate Bray-Curtis distances using OMU or individual metabolites.

```{r  dist_shared, R.options=list(max.print=15)}
bray_w_omus <- dist_shared(community_object = community_w_omus, size = 4000,
                          threshold = 100, diversity_index = "bray", subsample = TRUE)
bray_w_omus

bray_no_omus <- dist_shared(community_object = community_wo_omus, size = 4000,
                           threshold = 100, diversity_index = "bray", subsample = TRUE)
bray_no_omus
```

## Visualizations

There are a few different ways to visualize the data generated from this package. Below we have examples of a PCOA plot, Hierarchical Clustering graph, and a Network graph.

### PCOA

Principal Coordinate Analysis (PCOA) is a statistical technique used to show similarities between data. PCOA allows us, in particular, to use the distances generated from our ecological analysis. And, unlike Principle Component Analysis (PCA), it is not limited to only euclidean distances. So, we can use the distances generated from `dist_shared()` as input for this calculation. To start, we need to convert the `dist_shared()` result into cartisan coordinates. To do this, we can use the `stats::cmdscale()` function. Following, we can extract the points from the `stats::cmdscale()` function and rename them to be more readable. Finally, we plot the values. PCOA is a great way to look at your beta diversity results visually and shows us how similar samples are to each other.

```{r pcoa, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
dist_obj <- bray_w_omus

pcoa <- cmdscale(dist_obj, k = 2, eig = T, add = T)
pcoa_points <- as.data.frame(pcoa$points)
variance <- round(pcoa$eig*100/sum(pcoa$eig), 1)
names(pcoa_points) <- paste0("PCOA", seq(1:2))
pcoa_points$Samples <- rownames(pcoa_points)

meta_data <- get_meta_data(filtered_data)
for(i in 1:nrow(pcoa_points)) {
  idx <- which(pcoa_points$Samples[[i]] == meta_data$Injection)
  pcoa_points$Samples[[i]] <- meta_data$Biological_Group[[idx]]
}

pcoa_points |>
  ggplot(aes(x=PCOA1, y = PCOA2, color = Samples)) +
    geom_point(size = 5.5) +
    theme(
      legend.text = element_text(size = 10), 
      legend.title = element_text(size = 15)
    )
```

### Hierarchical Clustering

Another visualization approach is to perform hierarchical clustering using the `stats::hclust()` function. Hierarchical clustering generates a tree of samples based on their beta diversity values to show the similarities of the samples. The results are similar to that of PCoA but it uses a tree like structure to showcase relationships. It tends to be easier to look at relationships or clusters using this type of graph.

```{r hierarchical_clustering, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
hclust_result <- hclust(dist_obj, "complete")
par(cex=0.6, mar=c(5, 8, 4, 1))
plot(hclust_result)
```

### Network Plot

Sometimes, it is useful to see a visual map of your annotations. We can accomplish this using a network plot. By creating a network of data, we can see which compounds matched certain features. Its very useful for looking at relationships between compounds and features. And using `networkD3::simpleNetwork()` we can generate a very simple network plot.

```{r network_plot}
distance_df_annotations <- annotations[, c("query_ms2_id", "name", "score")]
simpleNetwork(distance_df_annotations, height="100px", width="100px", zoom = TRUE)
```


###  Using Group Averages
Another feature included inside of mums2, is the ability to run your analysis using group averages. Some researchers conduct mass spectrometry experiments by running samples in triplicates to account for noise. Since we know all three samples are the same, we can average out all of the samples into one group and work with one group. We are able to generate group averages by using the metadata file that was supplied when you imported your data. We can accomplish this by using the `convert_to_group_averages()` function. 

Below, you can see a copy of the functions we ran above. But this time, we are using group averages.  
```{r group_average}
  matched_avg <- convert_to_group_averages(matched_data = matched_data, mpactr_object = filtered_data)
  dist_avg <- dist_ms2(
    data = matched_avg, cutoff = 0.3, precursor_thresh = 2,
    score_params = modified_cosine_params(0.5), min_peaks = 0)

  cluster_results_avg <- cluster_data(
    distance_df = dist, ms2_match_data = matched_avg, cutoff = 0.3,
    cluster_method = "opticlust")

  annotations_avg <- annotate_ms2(mass_data = matched_avg, reference = reference_db, scoring_params = modified_cosine_params(0.5), 
                            ppm = 1000, min_score = 0.6, chemical_min_score = 0, cluster_data = cluster_results_avg)

  community_object_avg <- create_community_matrix_object(data = cluster_results_avg)
  head(get_community_matrix(community_object = community_object_avg), 3)
```

You can also see how the samples in the matched object have been adjusted based on the information supplied in the metadata file.
```{r samples}
  # Normal
  head(matched_data$samples)

  # Group Average
  head(matched_avg$samples)

  # The items in the Injection have been converted into the corresponding sample code.
  get_meta_data(filtered_data)[, c("Injection", "Sample_Code")]
```

### Combined data frames
mums2 also allows you to view your all of your generated data together using the `generate_a_combined_table()` function. This function will take your matched object (generated from `ms2_ms1_compare()`), your annotations object (generated from `annotate_ms2()`) and your cluster data (generated from `cluster_data()`) to generate a combined data.frame with all of the generated data. 
```{r combined_df}
  # For normal data
  head(generate_a_combined_table(matched_data = matched_data, annotations = annotations, cluster_data = cluster_results), 3)

  # For group averaged data
  head(generate_a_combined_table(matched_data = matched_avg, annotations = annotations_avg, cluster_data = cluster_results_avg), 3)


```


### Heat map
Another way to visualize diversity data is to use a heat map. We can easily see how correlated data is between samples with a heatmap.
```{r heat_map, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
dist_shared(community_object_avg, 4000, 100, "jaccard") |> 
  as.matrix() |>
  reshape2::melt(varnames = c("A", "B"), value.name = "distances") |>
  ggplot(aes(x = A, y = B, fill = distances)) +
  geom_tile() +
  scale_fill_gradient(low="#FF0000", high = "#FFFFFF")
```


### Box Plot
A box and whisker plot can also be used to depict the variation of the data. We can take advantage of this to look at the distributions inside of our generate beta and alpha diversity data. Below is a box plot using alpha diversity.
```{r box_plot, fig.width=10, fig.height=10, fig.fullwidth=TRUE}

alpha <- alpha_summary(community_object_avg, 4000, 100, "simpson") |>
  t() |> 
  as.data.frame()
alpha$name <- rownames(alpha)
alpha$group <- ""

# Convert Sample_Codes into Biological Groups
meta_data <- get_meta_data(filtered_data)
alpha$group <- sapply(alpha$name, function(x) {
   meta_data$Biological_Group[[which(x == meta_data$Sample_Code)[[1]]]]
}, USE.NAMES = FALSE)

# Plot 
colnames(alpha) <- c("distances", "sample_code", "group")
alpha |>
  ggplot(aes(x = sample_code, y = distances, group = group)) +
  geom_point()  + 
  geom_boxplot(alpha = 0.3) +
  facet_wrap(~group, scales = "free_x")



```


You can also look at the dissimilarities in the communities by using a box plot with beta diversity.
```{r box_plot_beta, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
dist_shared(community_object_avg, 4000, 100, "bray") |>
  as.matrix() |>
  reshape2::melt(varnames = c("A", "B"), value.name = "distances") |>
  ggplot(aes(x = A, y = distances, colour = A)) +
  geom_point(aes(color = B))  + 
  geom_boxplot(alpha = 0.1)

```
